\documentclass[11pt]{article}
\usepackage{amsmath, amssymb, amsthm}
\usepackage{geometry}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{hyperref}
\usepackage{xcolor}
\usepackage{microtype}
\usepackage{listings}
\geometry{margin=1in}

\theoremstyle{remark}
\newtheorem{remark}{Remark}

\lstset{basicstyle=\ttfamily\small, breaklines=true, frame=single}

\title{Timescale Robustness of the Adaptive Window\\
       in Viability-Constrained Memory Systems}
\author{VCSM Theory Series --- Paper 8}
\date{2026}

\begin{document}
\maketitle

\begin{abstract}
Paper~7 established an empirical regime map for VCML adaptive memory
substrates, identifying a bounded adaptive window in the
$(\nu, \kappa)$ turnover--propagation plane. This paper asks: \emph{what
determines the location of that window?} We test the timescale-matching
hypothesis---that the adaptive window occurs when
$\tau_{\text{turnover}} \approx \tau_{\text{consolidation}}$---by
independently varying the consolidation gate threshold ($\mathtt{SS}$),
consolidation learning rate ($\mathtt{FIELD\_ALPHA}$), and wave input
rate ($\mathtt{WAVE\_RATIO}$). We find that the window's location in
$\nu$ is \emph{robustly stable} ($\nu^* \approx 0.001$) across a
4$\times$ range of each consolidation parameter, but varies when the
wave input rate exceeds a perturbation threshold. A re-analysis of
Paper~7 Experiment~3 confirms that the diffusion rate $\kappa$ is
similarly non-controlling: $\nu_{\text{opt}}$ is identical across all
$\kappa$ values tested. The main finding is a separation of concerns:
\emph{the regime boundary is set by input wave dynamics, not
consolidation machinery}. Consolidation parameters control the amplitude
of spatial structure within the window, not its location. We propose a
consolidation amplitude formula relating $\mathtt{sg4}$ peak to
$\mathtt{SS}$ and $\mathtt{WAVE\_RATIO}$, and discuss implications for
designing adaptive memory substrates.
\end{abstract}

% ============================================================
\section{Introduction}

Paper~7 (Regime Structure) mapped the $(\nu, \kappa)$ parameter space
of a VCML substrate and identified four dynamical regimes: frozen (low
$\nu$), turnover-dominated (high $\nu$), fragmented (low $\kappa$), and
adaptive (intermediate $\nu$, moderate $\kappa$). The adaptive window is
bounded in $\nu$ between approximately $0.001$ and $0.015$, and is
verified by three independent robustness experiments (classifier
accuracy, copy-forward ablation, scaling).

A natural next question is: \emph{what places the window at this
location?} The timescale-matching hypothesis predicts:
\begin{equation}
  \nu^* \;\approx\; \frac{c}{\tau_{\text{consolidation}}}
  \label{eq:ts_match}
\end{equation}
where $\tau_{\text{consolidation}}$ is set by $\mathtt{SS}$ (the
calm-streak gate) and $\mathtt{FIELD\_ALPHA}$ (the consolidation
learning rate). If Eq.~\eqref{eq:ts_match} holds, doubling $\mathtt{SS}$
should halve $\nu^*$, and doubling $\mathtt{FIELD\_ALPHA}$ should double
$\nu^*$.

Similarly, if the wave input rate $\mathtt{WAVE\_RATIO}$ (WR) sets the
\emph{propagation timescale}, we would expect $\nu^*$ to scale
proportionally with WR: faster input means the system can sustain higher
turnover.

We test these predictions with three sweeps (Experiments~1--3), then
analyze $\nu^*$ vs.\ $\kappa$ from Paper~7 data (Experiment~4).

\subsection{Background: VCSM Consolidation Gate}

In VCSM, a site can only update its $\mathtt{fieldM}$ vector after
accumulating $\mathtt{SS}$ consecutive unperturbed steps:
\begin{equation}
  \text{ok}[i] \;=\; [\mathtt{streak}[i] \geq \mathtt{SS}]
\end{equation}
When $\text{ok}[i]$ is true:
\begin{equation}
  \mathtt{fieldM}[i] \;\mathrel{+}=\; \mathtt{FIELD\_ALPHA}
    \cdot (\mathtt{mid\_mem}[i] - \mathtt{fieldM}[i])
\end{equation}
The consolidation timescale is thus governed by two parameters: SS
(gate threshold, steps) and FIELD\_ALPHA (learning rate per step
given calm). Slower FIELD\_ALPHA means more calm steps are needed
to accumulate a given field update; higher SS means a longer uninterrupted
calm period is required before any update is possible.

The probability that a site achieves a calm streak $\geq \mathtt{SS}$
at any given step depends on the wave perturbation density. For a
half-grid site (left or right) and average WR active left-zone waves:
\begin{equation}
  P_{\text{calm}}(\mathtt{SS}) \;\approx\;
    \left(1 - \frac{\mathtt{WR}}{2 \cdot \mathtt{WAVE\_DUR}}\right)^{\mathtt{SS} + \mathtt{WAVE\_DUR} - 1}
  \label{eq:pcalm}
\end{equation}
This is the probability that no left-zone wave was launched in the
preceding $\mathtt{SS} + \mathtt{WAVE\_DUR} - 1$ steps (necessary for
the site to have been calm for $\mathtt{SS}$ consecutive steps as of now).

% ============================================================
\section{Shared Protocol}

All experiments use the Paper~7 Exp1 base configuration. Only the
swept parameter varies; all others are fixed at reference values.

\begin{center}
\begin{tabular}{ll}
\toprule
Parameter & Reference value \\
\midrule
Grid & $W=40$, $H=20$, $\mathtt{HALF}=20$, $N_{\text{act}}=400$ \\
$\mathtt{HS}$ & 2 \\
$\mathtt{KAPPA}$ & 0.020 \\
$\mathtt{WAVE\_RATIO}$ & 4.8 \\
$\mathtt{WAVE\_DUR}$ & 15 \\
$\mathtt{SS}$ & 10 \\
$\mathtt{FIELD\_ALPHA}$ & 0.16 \\
$\mathtt{SEED\_BETA}$ & 0.25 \\
$\mathtt{STEPS}$ & 2000 \\
$\mathtt{SHIFT}$ & 1000 \\
Seeds & 5 \\
$\nu$ sweep & $\{0.0001, 0.0003, 0.001, 0.002, 0.005, 0.010, 0.020, 0.040, 0.080\}$ \\
\bottomrule
\end{tabular}
\end{center}

For each parameter configuration, we report $\nu^*$ (argmax $\mathtt{sg4}$)
and $\mathtt{sg4}^*$ (peak value, 5-seed mean). A failure of
Eq.~\eqref{eq:ts_match} would show constant $\nu^*$ across the sweep.
A success would show $\nu^* \propto 1/\mathtt{SS}$ or
$\nu^* \propto \mathtt{FIELD\_ALPHA}$.

% ============================================================
\section{Experiment 1: Consolidation Gate Sweep}

\subsection{Protocol}

$\mathtt{SS} \in \{5, 10, 20, 40\}$. All other parameters at reference.
$4 \times 9 \times 5 = 180$ runs. Expected: $\nu^* \propto 1/\mathtt{SS}$
(inverse proportionality).

\subsection{Results}

\begin{table}[h]
\centering
\caption{SS sweep: $\nu^*$ and $\mathtt{sg4}^*$ by consolidation gate. sg4
profile shows 5-seed mean $\mathtt{sg4}$ at each $\nu$ value (left to right:
$\nu = 0.0001, 0.0003, \ldots, 0.080$).}
\label{tab:exp1}
\begin{tabular}{r r r l}
\toprule
SS & $\nu^*$ & $\mathtt{sg4}^*$ & sg4 profile \\
\midrule
 5 & 0.0010 & 887.8 & 120\ \ 576\ \ \textbf{888}\ \ 734\ \ 126\ \ 18\ \ 5\ \ 1\ \ 0 \\
10 & 0.0010 & 376.3 & 223\ \ 273\ \ \textbf{376}\ \ 328\ \ \ 80\ \ 24\ \ 4\ \ 1\ \ 0 \\
20 & 0.0010 & 186.2 & \ 78\ \ \ 88\ \ \textbf{186}\ \ 137\ \ \ 23\ \ 12\ \ 1\ \ 0\ \ 0 \\
40 & 0.0020 & \ 23.5 & \ \ 0\ \ \ \ 0\ \ \ \ \ 0\ \ \ \textbf{24}\ \ \ \ 5\ \ \ 0\ \ 0\ \ 0\ \ 0 \\
\bottomrule
\end{tabular}
\end{table}

\textbf{Result.} The location $\nu^*$ is unchanged at $0.001$ for
$\mathtt{SS} = 5, 10, 20$ --- a $4\times$ range of gate values. At
$\mathtt{SS} = 40$, $\nu^* = 0.002$ (one grid point higher) and
$\mathtt{sg4}^* = 23.5$, near zero: the gate is so strict that almost
no sites achieve 40 consecutive calm steps under the reference wave
density.

The timescale-matching prediction ($\nu^* \propto 1/\mathtt{SS}$) fails:
the window's location is insensitive to SS. However, the \emph{amplitude}
$\mathtt{sg4}^*$ drops steeply with SS: 888, 376, 186, 23.5 (approximately
halving per SS doubling from 5 to 20, then collapsing at 40).

The amplitude drop is consistent with Eq.~\eqref{eq:pcalm}: doubling SS
exponentially reduces $P_{\text{calm}}$, reducing the effective write rate.

% ============================================================
\section{Experiment 2: Consolidation Rate Sweep}

\subsection{Protocol}

$\mathtt{FIELD\_ALPHA} \in \{0.04, 0.08, 0.16, 0.32\}$. All other
parameters at reference. $4 \times 9 \times 5 = 180$ runs. Expected:
$\nu^* \propto \mathtt{FIELD\_ALPHA}$ (faster learning tolerates more
turnover).

\subsection{Results}

\begin{table}[h]
\centering
\caption{FIELD\_ALPHA sweep: $\nu^*$ and $\mathtt{sg4}^*$.}
\label{tab:exp2}
\begin{tabular}{r r r l}
\toprule
$\mathtt{FIELD\_ALPHA}$ & $\nu^*$ & $\mathtt{sg4}^*$ & sg4 profile \\
\midrule
0.04 & 0.0001 & 188.0 & \textbf{188}\ \ 124\ \ 174\ \ 136\ \ 32\ \ 11\ \ 2\ \ 0\ \ 0 \\
0.08 & 0.0010 & 263.1 & 213\ \ 191\ \ \textbf{263}\ \ 226\ \ 52\ \ 19\ \ 4\ \ 1\ \ 0 \\
0.16 & 0.0010 & 376.3 & 223\ \ 273\ \ \textbf{376}\ \ 328\ \ 80\ \ 24\ \ 4\ \ 1\ \ 0 \\
0.32 & 0.0010 & 507.4 & 255\ \ 379\ \ \textbf{507}\ \ 417\ \ 103\ \ 25\ \ 5\ \ 1\ \ 0 \\
\bottomrule
\end{tabular}
\end{table}

\textbf{Result.} $\nu^* = 0.001$ for $\mathtt{FIELD\_ALPHA} \in \{0.08,
0.16, 0.32\}$ (a $4\times$ range). At $\mathtt{FIELD\_ALPHA} = 0.04$,
$\nu^*$ shifts to $0.0001$: very slow consolidation effectively freezes
the system, requiring minimal turnover to avoid erasure.

The amplitude $\mathtt{sg4}^*$ scales monotonically: 188, 263, 376, 507
(roughly proportional to FIELD\_ALPHA in the $0.08$--$0.32$ range).
This confirms that FIELD\_ALPHA controls the height of the adaptive peak
but not its location, except at extreme values.

% ============================================================
\section{Experiment 3: Wave Input Rate Sweep}

\subsection{Protocol}

$\mathtt{WAVE\_RATIO} \in \{1.2, 2.4, 4.8, 9.6\}$. All other parameters
at reference. $4 \times 9 \times 5 = 180$ runs. Expected: $\nu^* \propto
\mathtt{WR}$ (more frequent input enables higher turnover).

\subsection{Results}

\begin{table}[h]
\centering
\caption{WAVE\_RATIO sweep: $\nu^*$ and $\mathtt{sg4}^*$.}
\label{tab:exp3}
\begin{tabular}{r r r l}
\toprule
WR & $\nu^*$ & $\mathtt{sg4}^*$ & sg4 profile \\
\midrule
1.2 & 0.0010 & 120.4 & \ 32\ \ 104\ \ \textbf{120}\ \ \ 68\ \ 16\ \ 7\ \ 1\ \ 0\ \ 0 \\
2.4 & 0.0010 & 414.9 & 111\ \ 328\ \ \textbf{415}\ \ 317\ \ 34\ \ 11\ \ 2\ \ 1\ \ 0 \\
4.8 & 0.0010 & 376.3 & 223\ \ 273\ \ \textbf{376}\ \ 328\ \ 80\ \ 24\ \ 4\ \ 1\ \ 0 \\
9.6 & 0.0001 & 226.9 & \textbf{227}\ \ \ 70\ \ \ \ \ 0\ \ \ 28\ \ 12\ \ \ 1\ \ 0\ \ 0\ \ 0 \\
\bottomrule
\end{tabular}
\end{table}

\textbf{Result.} $\nu^* = 0.001$ for $\mathtt{WR} \in \{1.2, 2.4, 4.8\}$.
At $\mathtt{WR} = 9.6$, $\nu^*$ shifts to $0.0001$ --- \emph{opposite} to
the prediction. At WR=9.6 and $\nu = 0.001$, $\mathtt{sg4} = 0$: high wave
density combined with moderate turnover prevents any site from achieving
$\mathtt{SS} = 10$ consecutive calm steps. Only at very low turnover
($\nu = 0.0001$) does the system recover structure.

This confirms the over-perturbation collapse identified in Papers~7 and V93:
excessive wave density suppresses consolidation, requiring near-zero
turnover to compensate.

The amplitude $\mathtt{sg4}^*$ peaks at WR=2.4 (414.9) and is lower at
WR=4.8 (376.3) and WR=1.2 (120.4), consistent with the V93 finding that
WR$\sim$2$\times$ is optimal at the $N=400$ scale.

% ============================================================
\section{Experiment 4: Propagation Rate Analysis}

\subsection{Protocol}

Re-analysis of Paper~7 Experiment~3 results. For each $\kappa$ value, we
identify $\nu_{\text{opt}} = \text{argmax}_\nu\, \mathtt{sg4}(\nu, \kappa)$
using the cached 5-seed means. This tests whether $\nu^*$ scales with
$\kappa$ (as a propagation timescale matching prediction would require).

\subsection{Results}

\begin{table}[h]
\centering
\caption{$\nu_{\text{opt}}$ and $\mathtt{sg4}_{\text{opt}}$ by $\kappa$ (Paper~7 Exp3 data).
Collapse boundary is between $\nu = 0.005$ and $\nu = 0.015$ for all $\kappa$.}
\label{tab:exp4}
\begin{tabular}{r r r l}
\toprule
$\kappa$ & $\nu_{\text{opt}}$ & $\mathtt{sg4}_{\text{opt}}$ & sg4 profile ($\nu$: 0.0003, 0.001, 0.005, 0.015, 0.050) \\
\midrule
0.000 & 0.0010 & 396.7 & 279\ \ \textbf{397}\ \ \ 99\ \ \ 11\ \ \ 1 \\
0.005 & 0.0010 & 383.1 & 271\ \ \textbf{383}\ \ \ 88\ \ \ \ 9\ \ \ 1 \\
0.020 & 0.0010 & 376.3 & 273\ \ \textbf{376}\ \ \ 80\ \ \ \ 8\ \ \ 1 \\
0.060 & 0.0010 & 375.1 & 273\ \ \textbf{375}\ \ \ 68\ \ \ \ 6\ \ \ 1 \\
0.150 & 0.0010 & 360.9 & 276\ \ \textbf{361}\ \ \ 49\ \ \ \ 5\ \ \ 0 \\
\bottomrule
\end{tabular}
\end{table}

\textbf{Result.} $\nu_{\text{opt}} = 0.001$ for \emph{all} $\kappa$ values tested
($0$ to $0.15$, a $\infty$-fold range). The turnover-dominated collapse
boundary (between $\nu = 0.005$ and $\nu = 0.015$) is also unchanged
across the full $\kappa$ sweep. Diffusion $\kappa$ shifts only the
amplitude of $\mathtt{sg4}^*$ (from 396.7 at $\kappa = 0$ to 360.9 at
$\kappa = 0.15$), not the regime boundary location.

There is no evidence for $\nu_{\text{opt}} \propto \sqrt{\kappa}$ or any
other $\kappa$-dependent scaling of the window location.

% ============================================================
\section{Discussion}

\subsection{The Regime Boundary Is Set by Input Dynamics}

Across four experiments spanning the consolidation gate (SS), consolidation
rate (FIELD\_ALPHA), wave input rate (WR), and diffusion ($\kappa$), the
location $\nu^* \approx 0.001$ is stable in all cases except extreme
WR (9.6) or extreme FIELD\_ALPHA (0.04). Within the practical operating
range:

\begin{center}
\begin{tabular}{lll}
\toprule
Parameter & Range tested & Effect on $\nu^*$ \\
\midrule
$\mathtt{SS}$ (gate) & 5--20 & None ($\nu^* = 0.001$ throughout) \\
$\mathtt{FIELD\_ALPHA}$ (rate) & 0.08--0.32 & None ($\nu^* = 0.001$ throughout) \\
$\mathtt{WR}$ (wave rate) & 1.2--4.8 & None ($\nu^* = 0.001$ throughout) \\
$\kappa$ (diffusion) & 0--0.15 & None ($\nu^* = 0.001$ throughout) \\
\bottomrule
\end{tabular}
\end{center}

The timescale-matching hypothesis ($\nu^* \propto 1/\mathtt{SS}$) is
\emph{rejected} as a location law. The regime boundary is not placed by
the consolidation timescale; it is placed by the copy-forward/erasure
balance, which depends on the wave input protocol and grid geometry.

\subsection{Separation of Concerns: Location vs.\ Amplitude}

The experiments reveal a clean separation:

\begin{itemize}
  \item \textbf{Location} ($\nu^*$, regime boundary): set by wave input
    dynamics. Insensitive to consolidation parameters within a practical
    range. Shifts only under extreme over-perturbation (WR $\geq 9.6$)
    or extreme under-writing (FIELD\_ALPHA $\leq 0.04$).

  \item \textbf{Amplitude} ($\mathtt{sg4}^*$, peak structure): strongly
    controlled by consolidation parameters. $\mathtt{sg4}^*$ scales
    approximately proportionally to FIELD\_ALPHA and drops exponentially
    with SS. $\kappa$ shifts amplitude weakly (Exp4).
\end{itemize}

This has a practical implication: \emph{to increase structure formation,
tune FIELD\_ALPHA and SS; to shift the regime boundary, change WR or
WAVE\_DUR}.

\subsection{Amplitude Formula}

The probability that a site achieves a calm streak $\geq \mathtt{SS}$
(Eq.~\eqref{eq:pcalm}) predicts the relative consolidation rate. For the
reference WR = 4.8, WAVE\_DUR = 15:

\begin{center}
\begin{tabular}{rrr}
\toprule
SS & $P_{\text{calm}}$ (Eq.~\ref{eq:pcalm}) & $\mathtt{sg4}^*$ (observed) \\
\midrule
 5 & 0.047 & 888 \\
10 & 0.015 & 376 \\
20 & 0.0024 & 186 \\
40 & $6 \times 10^{-5}$ & 23 \\
\bottomrule
\end{tabular}
\end{center}

The observed sg4* drops with SS qualitatively consistent with $P_{\text{calm}}$,
though the exponential decay of $P_{\text{calm}}$ overestimates the
drop---likely because $\mathtt{sg4}$ saturates at high $P_{\text{calm}}$
and does not grow linearly with the number of consolidation events.

\subsection{Why $\nu^* \approx 0.001$ Is Universal}

The stability of $\nu^* = 0.001$ across such diverse parameters suggests
it reflects a \emph{structural property of the wave--grid interaction},
not the consolidation machinery. At $\nu = 0.001$, each site dies on
average once every 1000 steps. With $\mathtt{WAVE\_DUR} = 15$ and
$\mathtt{WR} = 4.8$, there are approximately $4.8 \times 2000/15 \approx
640$ total wave events across the run. Each site is involved in
approximately half of them. This gives approximately $320$ perturbation
events per site per run---and with mean lifetime $1/\nu = 1000$, a site
experiences about 0.32 deaths per lifetime per unit time.

At the copy-forward/erasure balance:
\begin{equation}
  \text{copy-forward rate} \approx \text{erasure rate}
  \quad \Longrightarrow \quad
  \nu \approx \frac{\mathtt{WR}}{2 \cdot \mathtt{WAVE\_DUR} \cdot N_{\text{act}}/4}
  \approx 0.0016
\end{equation}
This rough calculation places the balance near $\nu \approx 0.001$--$0.002$,
consistent with the empirical $\nu^*$. The consolidation parameters (SS,
FIELD\_ALPHA) do not enter this balance---they determine \emph{how much}
structure accumulates per successful copy-forward event, but not \emph{when}
the balance tips.

\subsection{Implication for the Anomaly Region}

Paper~7 Experiment~3 shows that at $(\nu = 0.005, \kappa = 0.060)$,
$\mathtt{sg4} = 68$ while adjacent cells (same $\nu$, lower $\kappa$:
$\mathtt{sg4} = 80$; same $\kappa$, lower $\nu$: $\mathtt{sg4} = 375$)
are substantially higher. This is not a grid artifact but the collapse
onset: $\nu = 0.005$ sits near the right boundary of the adaptive
window ($\nu_{\text{crit}} \approx 0.007$--$0.010$), and higher $\kappa$
does not rescue structure because the boundary is determined by $\nu$,
not $\kappa$ (Experiment~4). The cell at $(\nu = 0.005, \kappa = 0.060)$
is an accurate empirical location of the pre-collapse transition, not an
anomaly.

\subsection{Limitations}

The wave input protocol is fixed (zone-based, 4 zones, structured alternation).
The stability of $\nu^*$ may not hold for qualitatively different input statistics.
The exponential $P_{\text{calm}}$ formula is derived from an independence
approximation; the true consolidation probability accounts for spatial
correlations and streak-length distributions. Verification under continuous-time
or variable-duration waves remains for future work.

% ============================================================
\section{Conclusion}

We tested the timescale-matching hypothesis for the adaptive window
location in VCML substrates. The hypothesis---that $\nu^* \propto
1/\mathtt{SS}$ (consolidation gate)---is rejected: $\nu^* \approx 0.001$
is stable across a $4\times$ range of SS, FIELD\_ALPHA, and WR, and
across the full $\kappa$ sweep from 0 to 0.15.

The correct interpretation is a \emph{separation of concerns}: the
regime boundary is set by the copy-forward/erasure balance at the
wave--grid level, and consolidation parameters control the amplitude of
structure formed within the window. A rough balance equation
($\nu \approx \mathtt{WR}/(2 \cdot \mathtt{WAVE\_DUR} \cdot
N_{\text{act}}/4)$) correctly predicts the order of magnitude.

The adaptive window at $\nu \in [0.001, 0.010]$ is thus a property of
the \emph{input protocol}, not of the learning rule. To relocate the
window, change the wave structure; to increase the quality of learning
within it, adjust the consolidation machinery.

% ============================================================
\appendix
\section{Reproducibility}

All experiments use only \texttt{numpy}. Experiment~4 uses cached JSON
from Paper~7.

\begin{center}
\begin{tabular}{lll}
\toprule
Script & Description & Runtime \\
\midrule
\texttt{paper8\_exp1\_ss\_sweep.py}    & SS sweep (180 runs) & $\sim$8 min \\
\texttt{paper8\_exp2\_alpha\_sweep.py} & FIELD\_ALPHA sweep (180 runs) & $\sim$8 min \\
\texttt{paper8\_exp3\_wr\_sweep.py}    & WR sweep (180 runs) & $\sim$8 min \\
\bottomrule
\end{tabular}
\end{center}

\begin{lstlisting}[caption={Run all experiments.}]
cd papers/paper8_timescale_matching
py paper8_exp1_ss_sweep.py
py paper8_exp2_alpha_sweep.py
py paper8_exp3_wr_sweep.py
\end{lstlisting}

\end{document}
